The Impact of Alternation

### 2017-05-04 draft.tex
ok

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\Sigma}$ consists of two distinct sorts
$\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
$\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
\in \ssorts{\sig}$. If $\#(f)=0$ then $f$ is a \emph{constant}. We
consider the constants $\mathbf{0}$ of sort $\Data$ and $\top,\bot$ of
sort $\Bool$, denoting the truth values true and false.

@@

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  \left(\Data^\I\right)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$. Given
formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
\eqiff^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
We omit mentioning the interpretation $\I$ when it is clear from the
context.

@@

This definition extends the classical notion of words from a finite
alphabet with a possibly infinite alphabet $\Sigma[\vec{x}]$. Clearly,
when $\Data^\I$ is sufficiently large or infinite, we can map the
elements of $\Sigma$ into designated elements of $\Data^\I$ and use a
special variable to encode the input events. However, keeping $\Sigma$
explicit in the following simplifies several technical points below,
without cluttering the presentation.

@@

In this section we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, does there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and $L(\overline{\A_1})
  = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclclc}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\
\overline{\phi} & \equiv & \phi & \text{ if $\phi \in Q$}\hspace*{1cm} &
\overline{\phi} & \equiv & \neg\phi & \text{if $\phi \not\in Q$ is an atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
equivalent, and in general:

@@

\begin{lemma}\label{lemma:closure}
  For any automata $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
  and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$ such that $Q_1
  \cap Q_2 = \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$,
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n) \in L(\A_\cup)$ be a word, where $n=0$ corresponds to
  the empty word. We prove by induction on $n\geq0$ that
  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n) \eqiff
  \Delta(\iota_1,a_1\ldots a_n) \vee \Delta(\iota_2,a_1\ldots
  a_n)$. The case $n=0$ follows from the definition of the initial
  configuration of $\A_\cup$. For the inductive step $n>0$,
  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$ is obtained from
  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by replacing each
  variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee \iota_2,a_1\ldots
    a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain: 
  \[\begin{array}{rcl}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iff & 
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee
  \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) \\
  & \iff & \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots
  a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{rcll}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & 
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
  & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
  \end{array}\]
  Let $\nu,\nu' : Q \cup \bigcup_{i=0}^n\vec{x}_i$ be two valuations
  such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x) = \mathbf{0}$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

@@

\section{Antichains and Abstraction Refinement for Emptiness}

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x = \overline{x} + k$, for $k \in \nat$ and $x \teq 0$,
with the classical interpretation of function symbols on
integers. Since reachability of a configuration $(q,\nu)$, where $q$
is a control state and $\nu$ a valuation of the variables, is
undecidable, the emptiness problem is undecidable for ADA, in general.

Consequently, we give up on the guarantee for termination and develop
semi-algorithmic methods that are supposed to be
both: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define $\Post{\A}(\phi,a) \equiv \exists
\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in
\Form^+(Q,\vec{x})$. We generalize $\Post{\A}$ to finite sequences of
input events: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
exist a sequence $u \in \Sigma^*$ such that the formula $\Accept{\A}(u)
\equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in F} \top \rightarrow
q \wedge \bigwedge_{q \in Q \setminus F} q \rightarrow \bot$ is
satisfiable ? A na\"ive semi-algorithm then enumerates all finite
sequences and checks for the satisfiability of $\Accept{\A}(u)$ for
each $u \in \Sigma^*$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need to
store $\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be
proved using $v$ alone. However, in practice, the semi-algorithm using
antichains diverges in a large number of cases, because, in many
interpretations, there exist infinite antichains\footnote{For
  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge
  x \teq 0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u_i) \mid
  u_1,u_2, \ldots \in \Sigma^*}$.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactitem}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item $\beta \cup \nu \models \Post{\A}(\iota,u) \Rightarrow \nu \in
  \Inv(\beta)$, and
%
\item $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu \not\models
  \Accept{\A}(u)$.
\end{compactitem}
Intuitively, a safety invariant maps every boolean valuation into a
set of data valuations, that contains the initial configuration (in
which all data variables are assumed to be $\mathbf{0}\in\Data^\I$),
over-approximates the set of reachable valuations and excludes those
valuations that satisfying the acceptance condition.

@@

Turning back to the issue of divergence of language emptiness
semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
stop at step $k$ as soon as $\bigvee_{i=1}^k \Post(\iota,u_i)$ defines
a safety invariant for $\A$. Although this condition can be
effectively checked using a decision procedure for the quantifier-free
first order theory $\theory(\sig,\I)$, of signature $\sig$ and
interpretation $\I$, there is no guarantee that this check will ever
succeed.

>>>>

A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
sorted \emph{function symbols}. To simplify the presentation, we
assume w.l.o.g. that $\ssorts{\sig}$ consists of two distinct sorts
$\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
$\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
\in \ssorts{\sig}$. If $\#(f)=0$ then $f$ is a \emph{constant}. We
consider the constants $\mathbf{0}$ of sort $\Data$ and $\top,\bot$ of
sort $\Bool$.

@@

An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
\item the sort $\Data$ into a non-empty set $\Data^\I$, 
%
\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
  $\top^\I = \true$, $\bot^\I = \false$, and
%
\item each function symbol $f$ into a total function $f^\I :
  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
  $\sigma(f)^I$ when $\#(f)=0$.
\end{inparaenum}
Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
each function symbol $f$ by its interpretation $f^\I$ and each
variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
$\I,\nu \models \phi$ if the formula obtained by replacing each term
$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
there exists a valuation $\nu$ such that $\I,\nu \models \phi$. The
\emph{theory} $\theory(\sig,\I)$ is the set of satisfiable formulae
written in the signature $\sig$, with the interpretation $\I$.

Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
\eqiff^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
We omit mentioning the interpretation $\I$ when it is clear from the
context.

@@

This definition generalizes the classical notion of words from a
finite alphabet to using a possibly infinite alphabet
$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
infinite, we can map the elements of $\Sigma$ into designated elements
of $\Data^\I$ and use a special variable to encode the input
events. However, keeping $\Sigma$ explicit in the following simplifies
several technical points below, without cluttering the presentation.

@@

In this section we tackle the following problems: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
  with the same set of variables $\vec{x}$, do there exist automata
  $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is it the case that
  $L(\A) = \emptyset$ ?
\end{compactenum}

@@

Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
negated occurrences of the boolean variables, we define the formula
$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
of $\phi$:
\[\begin{array}{lclclclclcl}
\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} && 
\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ is not an atom} \\
\overline{\phi} & \equiv & \phi \text{  if $\phi \in Q$}\hspace*{1cm} &&
\overline{\phi} & \equiv & \neg\phi \text{  if $\phi \not\in Q$ is an atom}
\end{array}\]
It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
equivalent, and in general:

@@

\begin{lemma}\label{lemma:closure}
  For any automata $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
  and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$ such that $Q_1
  \cap Q_2 = \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$,
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{rcl}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iff & 
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee
  \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) \\
  & \iff & \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots
  a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{rcll}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & 
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
  & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
  \end{array}\]
  Let $\nu,\nu' : Q \cup \bigcup_{i=0}^n\vec{x}_i$ be two valuations
  such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x) = \mathbf{0}$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

@@

\section{Antichains and Interpolants for Emptiness}

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x = \overline{x} + k$, for $k \in \nat$ and $x \teq 0$,
with the classical interpretation of function symbols on
integers. Since reachability of a control state is undecidable for
AVASS, the emptiness problem is undecidable for ADA.

Consequently, we give up on the termination guarantee and develop
semi-algorithmic methods that are supposed to be
both: \begin{compactenum}
\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
  the procedure will terminate and return a word $w \in L(\A)$), and
%
\item \emph{complete relative to termination}: if the procedure terminates
  without returning such a word, then $L(\A) = \emptyset$. 
\end{compactenum}

Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
(finite) input event alphabet is $\Sigma$, for the rest of this
section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
event $a \in \Sigma$, we define the \emph{post-image} function
$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
\Delta(\phi[\overline{\vec{x}}/\vec{x}], a)$, mapping each formula in
$\Form^+(Q,\vec{x})$ to a formula, also in $\Form^+(Q,\vec{x})$, that
defines the effect of firing a transition with input event $a$.

We generalize the post-image function to finite sequences of input
events, as follows: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
exist $u \in \Sigma^*$ such that the formula $\Accept{\A}(u) \equiv
\Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q
\rightarrow \bot)$ is satisfiable ? A na\"ive semi-algorithm
enumerates all finite sequences and checks the satisfiability of
$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
for the theory $\theory(\sig,\I)$.

Since no boolean variable from $Q$ occurs under negation in $\phi$, it
is easy to prove the following monotonicity property: given two
formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
\Sigma^*$. This suggest an improvement of the above semi-algorithm,
that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
  any $a,b \in A$.} w.r.t. the entailment partial order. This is
because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
$u$, since the non-emptiness of $\A$ can be proved using $v$
alone. However, in practice, the antichain based semi-algorithm
diverges in a large number of cases, since, in many interpretations,
there exist infinite antichains\footnote{For instance, when $\Data^\I
  = \nat$, an infinite antichain is $q \wedge x \teq 0, q \wedge x
  \teq 1, \ldots$}.

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation
$\nu : \vec{x} \mapsto \Data^\I$ of the data
variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
  \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
invariant. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration (in which all data variables are assumed to be
$\mathbf{0}\in\Data^\I$), over-approximates the set of reachable
valuations and excludes those valuations that satisfying the
acceptance condition.

@@

Turning back to the issue of divergence of language emptiness
semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
defines\footnote{ A formula $\phi(Q,\vec{x})$ defines $\Inv$ iff for
  all $\beta: Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow
  \Data^\I$: $\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. } a
safety invariant for $\A$. Although this condition can be effectively
checked using a decision procedure for the theory $\theory(\sig,\I)$,
there is no guarantee that this check will ever succeed.

The solution we adopt is the use of \emph{abstraction} to ensure
termination of invariant computations. However, it is worth pointing
out from the start that abstraction will allow us to build invariants
which are not necessarily safety invariants. To converge towards
meeting the latter condition, we shall resort to \emph{counterexample
  guided abstraction refinement} (CEGAR).

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can occur negated in some $\phi \in \preds$. We assume that $\preds$
is fixed for now, and refer to these formulae as \emph{predicates}.
Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
w.r.t. the predicates in $\preds$. The abstract version of the
post-image and acceptance formulae are defined as follows:
{\footnotesize\[
\abs{\Post{\A}}(\phi,\varepsilon) \equiv \phi 
\hspace*{2mm}
\abs{\Post{\A}}(\phi,ua) \equiv \abs{(\Post{\A}(\abs{\Post{\A}}(\phi,u),a))} 
\hspace*{2mm}
\abs{\Accept{\A}}(u) \equiv \abs{\Post{\A}}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
\]}
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \abs{\Post{\A}}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k>0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [1,k]$ such that
  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\abs{\Post{\A}}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k > 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [1,k]$ such that $\abs{\Post{\A}}(\iota,u) \eqiff
  \abs{\Post{\A}}(\iota,\mu(i))$, which concludes the proof. \qed}

We are left with fulfilling the second point from the definition of a
safety invariant. To this end, suppose that, for a given set $\preds$
of predicates, the invariant $\abs{\Inv}$, defined by the previous
lemma, meets point (\ref{it1:inv}) but not point (\ref{it2:inv}) from
the definition of invariants, where $\Post{\A}$ and $\Accept{\A}$
replace $\abs{\Post{\A}}$ and $\abs{\Accept{\A}}$, respectively. In
other words, there exists a finite sequence $u \in \Sigma^*$ such that
$\nu \in \abs{\Inv}(\beta)$ and $\beta\cup\nu \models
\abs{\Accept{\A}}(u)$, for some boolean $\beta : Q \rightarrow \booli$
and data $\nu : \vec{x} \rightarrow \Data^\I$ valuations. Such
a $u\in\Sigma^*$ is called a \emph{counterexample}.

Once a counterexample $u$ is discovered, there are two
possibilities. Either\begin{inparaenum}[(i)]
\item $\Accept{\A}(u)$ is satisfiable, in which case $L(\A) \neq \emptyset$, or
\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is called
  \emph{spurious}.
\end{inparaenum}
In the first case, our semi-algorithm stops and returns a witness for
non-emptiness, obtained from the satisfying valuation of
$\Accept{\A}(u)$ and in the second case, we must strenghten the
invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
that $\beta\cup\nu \models \abs{\Accept{\A}}(u)$. This strenghtening
is carried out by adding to $\preds$ predicates sufficient to exclude
the spurious counterexample.

In general, given an unsatisfiable conjunction $\pi \equiv
\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
  \in X}$ of any sort, an \emph{interpolant} for $\pi$ is a tuple
$\tuple{I_0(X), I_1(X), \ldots, I_n(X)}$ such
that:\begin{inparaenum}[(i)]
\item $I_0 \equiv \top$,  
%
\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
  I_i[X_i/X]$, for all $i \in [1,n]$, and
%
\item $I_n \equiv \bot$. 
\end{inparaenum}
A classical method for abstraction refinement is to add the elements
of the interpolant obtained from a proof of spuriousness to the set of
predicates. This guarantees progress, meaning that the particular
spurious counterexample, from which the interpolant was generated,
will never be revisited in the future. In many practical test cases,
this progress property eventually yields a safety invariant.
